---
title: \Huge\textbf {Preliminary Analysis}
author: |
        | Team Ninja
        | Department of Applied Mathematics & Statistics
        | State University of New York at Stony Brook
date: "2020/11/18"
output:
  pdf_document: default
  word_document: default
---


```{r}
rm(list=ls())
if (!require(readxl)) install.packages('readxl')
if (!require(ggplot2)) install.packages('ggplot2')
if (!require(devtools)) install.packages('devtools')
if (!require(covidcast)) install.packages('covidcast')
if (!require(dunn.test)) install.packages('dunn.test')
if (!require(ggpubr)) install.packages('ggpubr')
if (!require(openxlsx)) install.packages('openxlsx')
if (!require(hqreg)) install.packages('hqreg')

library(readxl)
library(ggplot2)
library(devtools)
library(covidcast)
library(dunn.test)
library(ggpubr)
library(openxlsx)
library(hqreg)
```

**Preprocessing**

```{r}
# Read data
mask = read.csv("mask-use-by-county.csv")
population = read_excel("co-est2019-annres.xlsx", skip = 4)[1:3142, c(1,13)]
rb = read_excel("rb_cont.xlsx")
tpolicy = read.csv("policy_date.csv")[1:51,]

colnames(population) = c("state", "population")
population$state = sub(".*,()","\\1",population$state)
population$state = trimws(population$state)
data = cbind(mask, population) # mask and population
data$total = ave(data$population, data$state, FUN = sum)
data$weight = data$population/data$total
data$mask = data$ALWAYS + data$FREQUENTLY # treat "always" and "frequently" as wearing mask
data$point = data$mask*data$weight
smask = tapply(data$point, data$state, FUN = sum) # population weighted
smask = as.data.frame(smask)[-9,]
data1 = cbind(smask, rb) # mask and rb
data1$population = unique(data$total)[-9]
data1$rep = data1$red - data1$blue
# Continuous variable of Red (vs. Blue)
data1$reps = "Swing"
data1$reps[(data1$red - data1$blue) > 0.1] = "Red"
data1$reps[(data1$blue - data1$red) > 0.1] = "Blue"
data1$reps = factor(data1$reps, levels = c("Red", "Swing", "Blue"), ordered = TRUE)
tpolicy = tpolicy[-9, 3:4]
for (i in 1:2){
  tpolicy[,i] = as.Date(tpolicy[,i], format = "%m/%d/%y")
}
data1 = cbind(data1, tpolicy)
```

**Box-plots and Regressions**

```{r}
my_comparisons <- list( c("Red", "Blue"), c("Red", "Swing"), c("Blue", "Swing") )
ggboxplot(data1, x = "reps", y = "smask",
          color = "reps", palette = "jco")+ 
  stat_compare_means(comparisons = my_comparisons, label.y = c(1, 0.9, 0.95))+
  stat_compare_means(label.y = 1.05) + 
  labs(x = "State Categories", y = "Mask wearing rates (%)", title = "Boxplot of Mask wearing rates of red, swing and blue states") + 
  scale_color_discrete(name = "Categories", labels = c("Red states","Swing states", "Blue states"))+ 
  theme(text = element_text(size = 13), axis.text.x = element_text(size = 13), axis.text.y = element_text(size = 13))
#ggsave("mask_rbs.png", width = 10, height = 7)
mean(data1$smask[data1$reps == "Red"])
mean(data1$smask[data1$reps == "Blue"])
shapiro.test(data1$smask)
wilcox.test(smask ~ reps, data = data1[data1$reps != "Swing", c("smask","reps")])
kruskal.test(smask ~ reps, data = data1)
dunn.test(data1$smask, data1$reps)

# Red vs. Blue (Continuous)
ggplot(data1, aes(x = rep, y = smask)) + geom_point() + geom_smooth(method=lm) + 
  labs(x = "Election rate of Republican - Election rate of Democrat", y = "Mask wearing rates", title = "Regression curve of mask wearing rate on election rate difference")+ theme(text = element_text(size = 12), axis.text.x = element_text(size = 12), axis.text.y = element_text(size = 12)) + annotate("text", x=0.2, y=0.95, label= "Pearson Correlation r = 0.698 (p-value = 4.48e-14)")
lm1 = lm(smask~rep, data1)
summary(lm1)

data1$stay_at_home_date = as.numeric(data1$stay_at_home_date) - as.numeric(as.Date("2020-01-01"))
data1$stay_at_home_expire_date = as.numeric(data1$stay_at_home_expire_date) - as.numeric(as.Date("2020-01-01"))
# Stay at home period
data1$home = data1$stay_at_home_expire_date - data1$stay_at_home_date
data1$home = jitter(data1$home)
ggboxplot(na.omit(data1), x = "reps", y = "home",
          color = "reps", palette = "jco")+ 
  stat_compare_means(comparisons = my_comparisons, label.y = c(100, 90, 95))+ 
  stat_compare_means(label.y = 105) + labs(x = "State Categories", y = "Stay at home days", title = "Boxplot of stay at home days of red, swing and blue states") + scale_color_discrete(name = "Categories", labels = c("Red states","Swing states", "Blue states"))+ theme(text = element_text(size = 13), axis.text.x = element_text(size = 13), axis.text.y = element_text(size = 13))

ggplot(na.omit(data1), aes(x = rep, y = home)) + geom_point() + geom_smooth(method=lm) + labs(x = "Election rate of Republican - Election rate of Democrat", y = "Stay at home days", title = "Regression curve of Stay at home days on election rate difference")+ theme(text = element_text(size = 12), axis.text.x = element_text(size = 12), axis.text.y = element_text(size = 12)) + annotate("text", x=0.2, y=75, label= "Pearson Correlation r = 0.16 (p-value = 0.0129)")
lm1 = lm(home~rep, data1)
summary(lm1)
```

**Time Series (Mask)**

```{r}
devtools::install_github("cmu-delphi/covidcast", ref = "main",subdir = "R-packages/covidcast")
cip_mask <- suppressMessages(
  covidcast_signal(data_source = "fb-survey", signal = "smoothed_wearing_mask", 
                   start_day = "2020-09-09", end_day = "2020-11-05",
                   geo_type = "state")
)
cip_mask = subset(cip_mask, select = c("geo_value","time_value","value"))
cip_mask <- tidyr::spread(cip_mask, geo_value, value)
mask_time = cip_mask[,2:52]
mask_time = unname(as.matrix(t(mask_time)))
mask_time = mask_time[c(2,1,4,3,5,6,7,9,10,11,12,14,15,16,13,17,18,19,22,21,20,23,24,26,25,27,30,34,31,32,33,35,28,29,36,37,38,39,40,41,42,43,44,45,47,46,48,50,49,51),] # adjust the order
population_state = unique(data$total)[-9]
sum1 = sum(population_state[data1$reps == "Red"])
sum2 = sum(population_state[data1$reps == "Blue"])
sum3 = sum(population_state[data1$reps == "Swing"])
weight_population = vector()
for(i in 1:50){
  if(data1$reps[i] == "Red"){
    weight_population[i] = population_state[i]/sum1
  }else if(data1$reps[i] == "Blue"){
    weight_population[i] = population_state[i]/sum2
  }else{
    weight_population[i] = population_state[i]/sum3
  }
}
red_prop = rep(0,58)
blue_prop = rep(0,58)
swing_prop = rep(0,58)
for(i in 1:50){
  if(data1$reps[i] == "Red"){
    red_prop = red_prop + mask_time[i,]*weight_population[i]
  }else if(data1$reps[i] == "Blue"){
    blue_prop = blue_prop + mask_time[i,]*weight_population[i]
  }else{
    swing_prop = swing_prop + mask_time[i,]*weight_population[i]
  }
}
red_prop = as.vector(as.matrix(red_prop))
blue_prop = as.vector(as.matrix(blue_prop))
swing_prop = as.vector(as.matrix(swing_prop))
date = seq(as.Date("2020-09-09"), as.Date("2020-11-05"), by="days")
mask_timeseries = as.data.frame(cbind(red_prop, blue_prop, swing_prop, date))
class(mask_timeseries$date) = "Date"

ggplot(mask_timeseries, aes(x = date)) + geom_line(aes(y=red_prop, colour = 'red')) + geom_line(aes(y=blue_prop, colour = 'blue')) + geom_line(aes(y=swing_prop, colour = 'green')) + labs(x = "Date", y = "Mask wearing rates (%)", title = "Time series of Mask wearing rates of red and blue states") + scale_color_discrete(name = "Categories", labels = c("Red states","Swing states", "Blue states"))+ theme(text = element_text(size = 12), axis.text.x = element_text(size = 12), axis.text.y = element_text(size = 13),legend.position = "none")
ggsave("mask_rbs_ts.png", width = 10, height = 7)
```

**Time Series(Daily confirmed proportion)**

```{r}
# Daily confirmed proportion
devtools::install_github("cmu-delphi/covidcast", ref = "main",
                         subdir = "R-packages/covidcast")
cip_state <- suppressMessages(
  covidcast_signal(data_source = "indicator-combination", signal = "confirmed_incidence_prop", 
                   start_day = "2020-04-06", end_day = "2020-11-05",
                   geo_type = "state")
)
cip_state <- subset(cip_state, select = c("geo_value","time_value","value"))
cip_state <- tidyr::spread(cip_state, geo_value, value)
confirmed_prop = cip_state[,2:53]
confirmed_prop = unname(as.matrix(t(confirmed_prop)))
confirmed_prop = confirmed_prop[c(2,1,4,3,5,6,7,9,10,11,12,14,15,16,13,17,18,19,22,21,20,23,24,26,25,27,30,34,31,32,33,35,28,29,36,37,38,39,41,42,43,44,45,46,48,47,49,51,50,52),] # adjust the order
population_state = unique(data$total)[-9]
sum1 = sum(population_state[data1$reps == "Red"])
sum2 = sum(population_state[data1$reps == "Blue"])
sum3 = sum(population_state[data1$reps == "Swing"])
weight_population = vector()
for(i in 1:50){
  if(data1$reps[i] == "Red"){
    weight_population[i] = population_state[i]/sum1
  }else if(data1$reps[i] == "Blue"){
    weight_population[i] = population_state[i]/sum2
  }else{
    weight_population[i] = population_state[i]/sum3
  }
}
red_prop = rep(0,214)
blue_prop = rep(0,214)
swing_prop = rep(0,214)
for(i in 1:50){
  if(data1$reps[i] == "Red"){
    red_prop = red_prop + confirmed_prop[i,]*weight_population[i]
  }else if(data1$reps[i] == "Blue"){
    blue_prop = blue_prop + confirmed_prop[i,]*weight_population[i]
  }else{
    swing_prop = swing_prop + confirmed_prop[i,]*weight_population[i]
  }
}
red_prop = as.vector(as.matrix(red_prop))
blue_prop = as.vector(as.matrix(blue_prop))
swing_prop = as.vector(as.matrix(swing_prop))
date = seq(as.Date("2020-04-06"), as.Date("2020-11-05"), by="days")
confirmed_time_prop = as.data.frame(cbind(red_prop, blue_prop, swing_prop, date))
class(confirmed_time_prop$date) = "Date"

ggplot(confirmed_time_prop, aes(x = date)) + geom_line(aes(y=blue_prop, colour = 'red')) + geom_line(aes(y=red_prop, colour = 'blue')) + geom_line(aes(y=swing_prop, colour = 'green')) + labs(x = "Date", y = "Population weighted daily confirmed proportion (%)", title = "Time series of weighted daily confirmed proportion of red and blue states") + scale_color_discrete(name = "Categories", labels = c("Red states","Swing states", "Blue states"))+ theme(text = element_text(size = 10), axis.text.x = element_text(size = 10), axis.text.y = element_text(size = 10),legend.position = c(.2, .95),
    legend.justification = c("right", "top"),
    legend.box.just = "right",
    legend.margin = margin(6, 6, 6, 6)) + scale_x_date(breaks = "1 month", date_labels = "%b")
```

** Correlation**
## Step 1: Daily Cases Data Manipulation

Download the 7-day average daily cases between 2020-7-2 to 2020-10-31.

```{r datamani}
devtools::install_github("cmu-delphi/covidcast", ref = "main",
                         subdir = "R-packages/covidcast")
Date <- seq.Date(from = as.Date("2020/07/02",format = "%Y/%m/%d"), by = "day", 
                 length.out = 122)

sevendaysaverage<- suppressMessages(
  covidcast_signal(data_source = "indicator-combination", 
                   signal = "confirmed_7dav_incidence_prop", 
                   start_day = "2020-07-2", end_day = "2020-10-31",
                   geo_type = "state")
)

A<-matrix(nrow=52,ncol=123)
dfsevenaverage<-(A)
colnames(dfsevenaverage)[2:123]<-as.character(Date)
colnames(dfsevenaverage)[1]<-'state'
dfsevenaverage[1:52,1]<-sevendaysaverage[1:52,3]
for (i in 1:52){
  for(j in 1:122){
    dfsevenaverage[i,j+1]=sevendaysaverage[i+(j-1)*52,7]
    
  }
  
}

dfseven50<-data.frame(dfsevenaverage[-c(8,40),])
print(head(dfseven50[,1:5]))

```

## Step 2: Read Mask Data

Read the mask wearing percantage from local. It is got from NYtimes.
https://www.nytimes.com/interactive/2020/07/17/upshot/coronavirus-face-mask-map.html

The mask wearing percentage for all states was collected between July 2 and July 14

```{r mask}
covidfeatures<-read.xlsx('Mask wearing percentage.xlsx')
```

## Step 3 Normalize the Mask Wearing Percantage and Numeric the Daily Cases

Normalize the mask wearing percantage and numeric the daily cases

```{r}
covidlableuse<-as.matrix(dfseven50[2:length(dfseven50)])
covidfeaturesuse<-as.matrix(covidfeatures[2:length(covidfeatures)])

for(i in 1:dim(covidfeaturesuse)[2]){
  covidfeaturesuse[,i]=scale(covidfeaturesuse[,i],center=TRUE,scale=TRUE)
}

covidlableuse2<-matrix(nrow=50,ncol=122)

for (i in 1:50){
  covidlableuse2[i,]<-as.numeric(covidlableuse[i,])
}

```

## Step 4 Get the Correlations and Draw the Correlations

Get the Correlations and Draw the Correlations
The mask wearing percentage for all states was collected between July 2 and July 14 (in green shadow)

```{r}

correlationvector<-c()

for(i in 1:122){correlationvector<-c(correlationvector,cor(covidlableuse2[,i],covidfeaturesuse[,1]))}


df<-data.frame(Date=Date,Correlation=correlationvector)

shadow <- data.frame (xmin=Date[1], xmax=Date[13], ymin=-Inf, ymax=Inf)


ggplot(data = df, mapping = aes(x = Date, y = Correlation, group = 1)) + geom_line() + 
  xlab('Date')+
  labs(title = "Correlation between the mask wearing rate and daily confirmed cases")+
  geom_rect(data=shadow, aes(xmin=xmin, xmax=xmax, ymin=ymin, ymax=ymax), 
            fill="green", alpha=0.1, inherit.aes = FALSE)
```
